A Framework for Reproducible Data Science using Nix

### 2025-10-17 paper.tex
getting ready

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  article]{jss}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
\makeatother
\makeatletter
\makeatother
\makeatletter
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Reproducible development environments with rix},
  pdfauthor={Bruno Rodrigues; Philipp Baumann},
  pdfkeywords={reproducibility, R, Nix},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%% Author information
\author{Bruno Rodrigues~\orcidlink{0000-0002-3211-3689}\\Ministry of
Research and Higher education, Luxembourg \And Philipp Baumann\\Plus
Affiliation}
\Plainauthor{Bruno Rodrigues, Philipp Baumann} %% comma-separated

\title{Reproducible development environments with rix}
\Plaintitle{Reproducible development environments with
rix} %% without formatting

%% an abstract and keywords
\Abstract{In order to create an analysis that is easily reproducible, it
is not enough to write clean code and document it well. One must also
make sure to list all the dependencies of the analysis clearly and
ideally provide an easy way to install said dependencies. There are
several tools that can be used to list dependencies and to make them
easily installable by someone that wishes to reproduce a study, such as
\proglang{Docker}, a containerization solution. This paper will present
the Nix package manager, and an \proglang{R} package called \pkg{rix}
that lowers Nix's learning curve for users of the \proglang{R}
programming language.}

%% at least one keyword must be supplied
\Keywords{reproducibility, \proglang{R}, \proglang{Nix}}
\Plainkeywords{reproducibility, R, Nix}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Bruno Rodrigues\\
Department of Statistics\\
18, Montée de la Pétrusse\\
Luxembourg Luxembourg\\
E-mail: \email{bruno@brodrigues.co}\\
URL: \url{https://www.brodrigues.co}\\
\\~
Philipp Baumann\\
\\~

@@

\begin{document}
\maketitle
\section{Introduction: Reproducibility is also about
software}\label{sec-intro}

\citet{peng2011} introduced the idea of reproducibility being on a
continuum: on one of the ends of this continuum, we only have access to
the paper describing the studies, which is not reproducible at all.
Then, if in addition, to this paper we make the original source code of
the analysis that was written to compute the results of the study
available, reproducibility is improved, albeit only by a little. Adding
the original data improves reproducibility yet again. Finally, if to all
this we add what Roger Peng named the \emph{linked and executable code
and data}, we reach the gold standard of full replication.

What is this \emph{linked and executable code and data}? Another way to
name this crucial piece of the reproducibility puzzle is
\emph{computational environment}. The computational environment is all
the software required to actually run the analysis. Here too, we can
speak of a continuum. One could simply name and list the software used:
for example, the \proglang{R} programming language. Sometimes, authors
have the courtesy to also state the version of \proglang{R} used. Some
authors go further, and also list the packages used, and ideally with
their versions as well. Authors rarely state the operating system on
which the analysis was done, even though it has been shown that running
the same analysis with the same software but on different operating
systems could lead to different results, as described in
\citet{neupane2019}. Authors also only very rarely provide instructions
to install the required tools and software in order to reproduce their
studies.

Tools can be used at each of these steps to reach the gold standard of
full replication. Let's first consider the task of listing the software
used. \proglang{R} provides the \texttt{sessionInfo()} function whose
output can be saved into a file. Below is an example output of
\texttt{sessionInfo()}:

\begin{verbatim}
sessionInfo()
\end{verbatim}

\begin{verbatim}
R version 4.3.2 (2023-10-31)
Platform: aarch64-unknown-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

@@

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods  
[7] base

other attached packages:
[1] nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163

loaded via a namespace (and not attached):
[1] compiler_4.3.2 Matrix_1.6-1.1 tools_4.3.2    splines_4.3.2 
[5] grid_4.3.2     lattice_0.21-9
\end{verbatim}

If an author provides this information, other people trying to reproduce
the study (or the author him- or herself in the future) can read this
file and see which version of \proglang{R} was used, and which packages
(and their versions) were used as well. However, others would still need
to install the correct software themselves. An alternative to this is
instead to use the \pkg{renv} package which generates a so-called
\texttt{renv.lock} file which also lists \proglang{R} and package
versions. Here is an example of such an \texttt{renv.lock} file:

\begin{verbatim}
{
"R": {
  "Version": "4.2.2",
  "Repositories": [
  {
   "Name": "CRAN",
   "URL": "https://packagemanager.rstudio.com/all/latest"
  }
  ]
},
"Packages": {
  "MASS": {
    "Package": "MASS",
    "Version": "7.3-58.1",
    "Source": "Repository",
    "Repository": "CRAN",
    "Hash": "762e1804143a332333c054759f89a706",
    "Requirements": []
  },
  "Matrix": {
    "Package": "Matrix",
    "Version": "1.5-1",
    "Source": "Repository",
    "Repository": "CRAN",
    "Hash": "539dc0c0c05636812f1080f473d2c177",
    "Requirements": [
      "lattice"
    ]

    ***and many more packages***
\end{verbatim}

This file lists every package alongside their versions and the
repository from which they were downloaded. Generating this file only
requires one to run the \texttt{renv::init()} function. Someone else can
then restore the same package library by running
\texttt{renv::restore()}. The exact same packages get installed in an
isolated, project-specific, library which doesn't interfere with the
other, main, library of the user. \pkg{renv} does not restore the
\proglang{R} itself though, so installing the right version of R needs
to be handled separately. Before continuing, it should be noted that
other packages exist which provide similar functionality to \pkg{renv}:
there is \pkg{groundhog} by \citet{simonsohn2023} which makes it rather
easy to install packages as they were on CRAN at a given date. For
example, the code snippet below install the \pkg{purrr} and
\pkg{ggplot2} packages as they were on April 4th, 2017:

\begin{verbatim}
groundhog.library("
    library(purrr)
    library(ggplot2)",
    "2017-10-04",
    tolerate.R.version = "4.2.2")
\end{verbatim}

These packages also get installed in a project-specfic library so there
is no interferences between these packages and other versions of the
same packages that one might use for other projects. Because
\pkg{groundhog} does not install \proglang{R} itself, users should
either install the required version themselves, or they should use the
\texttt{tolerate.R.version} argument as shown in the example above.
Otherwise, \pkg{groundhog} would not continue with the installation of
the packages. Another such package, developed by \citet{chan2023} is
\pkg{rang}, which also installs packages as they were on a given date.
Yet another way to install packages as they were on a give date is to
use the Posit Package Manager, which provides snapshots of CRAN. For
example, to install the required packages for an analysis as they were
on the 30th of June 2023, one could add the following line to the
\texttt{.Rprofile} file:

\begin{verbatim}
options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
                               __linux__/jammy/2023-06-30"))
\end{verbatim}

The \texttt{.Rprofile} file gets read by R when starting a new session,
which means that every call to the \texttt{install.packages()} function
will now install the packages from this snapshotted mirror.

The next step in reaching the gold standard of reproducibility would be
to not only install the right packages used for the analysis, but also
the right version of \proglang{R}. Of course, it would be possible to
install the right version manually, but here too, there are tools that
simplify the process such as \proglang{rig} by the R infrastructure team
\citeyearpar{rlib2023}. One could thus install the right version of
\proglang{R} using \proglang{rig} allows to easily install different
versions of \proglang{R}, one could use to install the right version to
reproduce an analysis, and then use one of the listed packages above to
install the right library of of \proglang{R} packages. However, this
involves many manual steps and is thus error prone. It is also
time-consuming to do so, especially if one wants to reproduce many
studies.

The final step towards the gold standard of reproducibility is to
package the right version of \proglang{R} and \proglang{R} packages
inside a \proglang{Docker} image. \proglang{Docker} is a
containerisation tool: using \proglang{Docker} it is possible to package
some \emph{data product} with its dependencies into an image. From this
image, containers can be run to reproduce the data product with its
exact dependencies. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that comes with
many software dependencies. \emph{Dockerizing} an analysis consists in
first building an image: in the build step, the dependencies of the
analysis have to be installed and this can be achieved using the tools
mentioned above. The scripts to run the analysis and the data are also
added to the image at build time. Instances of that image, called
containers, can then be executed, which give access to the exact
development environment originally used for the study. It is also
possible to script the building of the analysis itself, such that in one
single call to the \texttt{docker\ run} command, a complete analysis is
reproduced.

The Rocker project initiated by \citet{boettiger2017} provides many
pre-built \proglang{Docker} to the \proglang{R} community of users.
These images can be used as bases to build other images containing
statistical analyses more easily than starting from a bare-bones image.

An optional step towards the gold standard is to use a build automation
tool such as \proglang{Make} to run the whole analysis when the
container is executed. Build automation tools make it easier to run
arbitrary code in a series of well-defined steps. \proglang{R}
programmers can use the \pkg{targets} by \citet{landau2021} as a build
automation tool.

\citet{mcdermott2021} is an example of a scientific study that reached
the gold standard of reproducibility. The author of this paper set up an
accompagnying Github repository to the paper\footnote{https://github.com/grantmcdermott/skeptic-priors}
containing all the instructions to install the required software and
then run the analysis. If we take a closer look at this repository, we
will notice that many of the tools previously mentioned were used to
capture the compatutational environment and make it available to other
researchers:

\begin{itemize}
\tightlist
\item
  Packages and their versions were listed and saved into an
  \texttt{renv.lock} file;
\item
  A \texttt{Makefile} was used to run the whole analysis and compile the
  paper;
\item
  A \texttt{Dockerfile} was used to provide the complete computational
  environment, including the right version of \proglang{R} and run the
  whole analysis easily.
\end{itemize}

However, reaching this gold standard is quite costly: one needs to learn
a tool to deal with package versions, then \proglang{Docker} for all the
other software, including the programming language that was used.
Ideally, a build automation tool such as \proglang{make} should be added
to the list. It should also be noted that these requirements are not
limited to the \proglang{R} programming language. A very similar
approach should be taken if one uses \proglang{Python} for statistical
analysis instead.

As an alternative to the above approach, I will present the
\proglang{Nix} package manager, which available for all major operating
system and focuses on installing and building software in a reproducible
manner. To make \proglang{Nix} more accessible to \proglang{R}
programmers, I developed the \pkg{rix} which I will also present in this
article.

\section{The Nix package manager}\label{the-nix-package-manager}

\proglang{Nix} is a package manager that can be used to install and
build software in a completely reproducible manner. As of writing, it
contains more than 80.000 packages, and the entirety of CRAN and
Bioconductor is available through \proglang{Nix}'s repositories. This
means that using \proglang{Nix}, it is possible to install not only
\proglang{R}, but also all the packages required for a project. The
reason why one should use \proglang{Nix} to install \proglang{R}
packages, and not use the usual, built-in, \texttt{install.packages()}
\proglang{R} function instead, is that when installing a package with
\proglang{Nix}, \proglang{Nix} makes sure to install every dependency of
every package, whether this dependency is another \proglang{R} package
or a system-level dependency.

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  article]{jss}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
\makeatother
\makeatletter
\makeatother
\makeatletter
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A Framework for Reproducible Data Science using Nix},
  pdfauthor={Bruno Rodrigues},
  pdfkeywords={reproducibility, R, Nix},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%% Author information
\author{Bruno Rodrigues~\orcidlink{0000-0002-3211-3689}\\Ministry of
Research and Higher education, Luxembourg}
\Plainauthor{Bruno Rodrigues} %% comma-separated

\title{A Framework for Reproducible Data Science using Nix}
\Plaintitle{A Framework for Reproducible Data Science using
Nix} %% without formatting

%% an abstract and keywords
\Abstract{To create a reproducible analysis, it is not enough to write
clean, well-documented and tested code. One must also ensure that all
software dependencies are clearly declared and, ideally, provide a
simple way to install them. Several tools exist for this purpose, such
as the containerization solution \pkg{Docker}. A complete solution,
however, requires not only managing the computational environment but
also orchestrating the execution of the analysis itself in a robust and
automated way. This paper presents a comprehensive framework based on
the \pkg{Nix} package manager to solve both problems. We first introduce
the \proglang{R} package \pkg{rix}, which simplifies the creation of
reproducible environments with \pkg{Nix}. We then introduce
\pkg{rixpress}, a second \proglang{R} package that leverages these
environments to define and execute complex, multi-language analytical
pipelines. Together, these tools provide a powerful, unified approach to
achieving high standards of computational reproducibility.}

%% at least one keyword must be supplied
\Keywords{reproducibility, \proglang{R}, \pkg{Nix}}
\Plainkeywords{reproducibility, R, Nix}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Bruno Rodrigues\\
Department of Statistics\\
18, Montée de la Pétrusse\\
Luxembourg Luxembourg\\
E-mail: \email{bruno@brodrigues.co}\\
URL: \url{https://www.brodrigues.co}\\
\\~

@@

\begin{document}
\maketitle

\section{Introduction: Reproducibility is also about
software}\label{sec-intro}

\citet{peng2011} introduced the concept of reproducibility as a
\emph{continuum}. At one end lies the least reproducible state, where
only a paper describing the study is available. Reproducibility improves
when authors share the original source code, improves further when they
include the underlying data, and reaches its highest level when what
Roger Peng called \emph{linked and executable code and data} are
provided.

By \emph{linked and executable code and data}, Peng referred to compiled
source code and runnable scripts. In this paper, we interpret this
notion more broadly as the \emph{computational environment}: the
complete set of software required to execute an analysis. Here too, a
continuum exists. At the minimal end, authors might only name the main
software used---say, the \proglang{R} programming language. More careful
authors might also specify the version of \proglang{R}, or list the
additional packages and their versions. Rarely, however, do authors
specify the operating system on which the analysis was performed, even
though differences in operating systems can lead to divergent results
when using the same code and software versions, as shown by
\citet{neupane2019}. It is even less common for authors to provide
step-by-step installation instructions for the required software stack.

Even when such instructions are given, they often fail across different
platforms or versions of the same platform. This lack of portability not
only hinders reproducibility but also complicates everyday research
workflows. Researchers working across multiple machines must be able to
recreate the same environment consistently, and collaborators must share
identical computational setups to avoid inconsistencies.

Finally, once the execution environment is correctly configured,
additional clarity is needed on how to \emph{run} the project itself.
Which packages should be loaded first? Which scripts should be executed,
and in what order? Without clear documentation or a logically organized
project structure, these operational details become yet another barrier
to reproducibility.

A range of tools now exist to help researchers approach the gold
standard of full reproducibility, or to consistently deploy the same
development environment across multiple machines. Let us first consider
the most basic step in this process: listing the software used. In
\proglang{R}, the \texttt{sessionInfo()} function provides a concise
summary of the software environment, including the R version, platform
details, and all loaded packages. Its output can be saved to a file and
included as part of a study's reproducibility record. Below is an
example output from \texttt{sessionInfo()}:

\begin{CodeInput}
R> sessionInfo()
\end{CodeInput}
\begin{CodeOutput}
R version 4.3.2 (2023-10-31)
Platform: aarch64-unknown-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

@@

attached base packages:
 stats     graphics  grDevices utils     datasets  methods   base

other attached packages:
 nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163

loaded via a namespace (and not attached):
 compiler_4.3.2  Matrix_1.6-1.1  tools_4.3.2     splines_4.3.2
 grid_4.3.2      lattice_0.21-9 
\end{CodeOutput}

When an author includes this information, others attempting to reproduce
the study (future readers, collaborators, or even the author at a later
time) can easily see which version of \proglang{R} and which packages
(with their versions) were used. However, reproducing the environment
still requires manually installing the correct package versions. This
can be difficult, especially when packages depend on system-level
libraries. For example, the \pkg{nloptr} package requires a precompiled
\proglang{nlopt} binary, or alternatively, \proglang{cmake} when
building from source on Linux or macOS.

A more robust approach than simply listing package versions is to use
the \pkg{renv} package. \pkg{renv} captures the project's software state
and writes it to a \texttt{renv.lock} file, which includes the exact
versions of \proglang{R} and all required packages. This lockfile serves
as a blueprint for restoring the environment automatically, ensuring
that others can recreate the same setup with minimal effort. Below is an
example of an \texttt{renv.lock} file:

\begin{Code}
{
  "R": {
    "Version": "4.2.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.rstudio.com/all/latest"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-58.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "762e1804143a332333c054759f89a706",
      "Requirements": []
    },
    "Matrix": {
      "Package": "Matrix",
      "Version": "1.5-1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "539dc0c0c05636812f1080f473d2c177",
      "Requirements": [
        "lattice"
      ]
    }

    ... lines below omitted ...
  }
}
\end{Code}

This lockfile lists each package alongside its version and the
repository from which it was downloaded. Creating it requires only a
single command: \texttt{renv::init()}. Others can then reproduce the
same package library by running \texttt{renv::restore()}, which installs
the exact package versions in an isolated, project-specific library that
does not interfere with the user's global R library.

However, \pkg{renv} does not restore the version of \proglang{R} itself.
Installing the correct R version must therefore be done separately.
Similarly, \texttt{renv} does not handle system-level dependencies such
as \proglang{cmake}, which is required by \pkg{nloptr} on Linux and
macOS. These dependencies must be managed manually or through other
tools. Further details are discussed in the \pkg{renv} documentation:
\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

Before continuing, it should be noted that other packages exist which
provide similar functionality to \pkg{renv}: there is \pkg{groundhog} by
\citet{simonsohn2023} which makes it rather easy to install packages as
they were on CRAN at a given date. For example, the code snippet below
installs the \pkg{purrr} and \pkg{ggplot2} packages as they were on
April 4th, 2017:

\begin{CodeInput}
R> groundhog.library("
+   library('purrr')
+   library('ggplot2')",
+   "2017-10-04",
+   tolerate.R.version = "4.2.2")
\end{CodeInput}

Packages installed with \pkg{groundhog} are placed in a project-specific
library, ensuring they do not interfere with other versions of the same
packages used in different projects. Since \pkg{groundhog} does not
install \proglang{R} itself, users must either install the required R
version manually or use the \texttt{tolerate.R.version} argument, as
illustrated in the example above. Without this argument, \pkg{groundhog}
will not proceed with package installation if the R version does not
match the expected one.

An alternative to \pkg{renv} is \pkg{rang}, developed by
\citet{chan2023}, which similarly allows users to install packages as
they existed on a specific date. Another approach is to use the Posit
Package Manager, which provides dated snapshots of CRAN. For instance,
to ensure that packages are installed as they were on June 30, 2023, one
could include the following line in the \texttt{.Rprofile} file:

\begin{CodeInput}
R> options(repos =
+    c(REPO_NAME =
+      "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
+    )
+  )
\end{CodeInput}

The \texttt{.Rprofile} file is read by \proglang{R} at the start of each
new session, meaning that every call to \texttt{install.packages()} will
install packages from the specified snapshot mirror. However, unless
users explicitly manage separate libraries for different projects, the
Posit Package Manager will install all packages for all projects as they
existed on that snapshot date.

The next step toward the gold standard of reproducibility is ensuring
not only that the correct package versions are installed, but also that
the correct version of \proglang{R} itself is used. While this can be
done manually, dedicated tools make the process more efficient. One such
tool is \pkg{rig}, developed by the \proglang{R} Infrastructure Team
\citeyearpar{rlib2023}. \pkg{rig} simplifies installing and managing
multiple R versions, allowing users to match the exact version needed
for a given analysis. Once the correct R version is installed, one can
then use tools such as \pkg{renv}, \pkg{groundhog}, or \pkg{rang} to
recreate the associated package library.

However, this multi-step process remains manual, error-prone, and
time-consuming, highlighting the need for more comprehensive and
automated solutions to achieve full computational reproducibility.

The final tool for achieving the gold standard of reproducibility is to
bundle the exact version of \proglang{R} and its packages within a
\pkg{Docker} image. \pkg{Docker} is a containerization platform that
allows packaging a \emph{data product} together with all its
dependencies into a self-contained image. A statistical analysis can be
viewed as such a data product, requiring a specific set of software
dependencies to run. \emph{Dockerizing} an analysis involves building an
image that installs these dependencies, often using the tools mentioned
earlier, and includes the analysis scripts and data (but data can also
be dynamically made available at run-time, to avoid having to copy it
inside the image). The steps to create this image are specified in a
simple text file called a \texttt{Dockerfile}, which defines the
structure and content of the \pkg{Docker} image.

Once the image is built, the analysis can be executed inside a
\emph{container}, which is a running instance of the image. Containers
are typically run non-interactively, allowing the complete computational
environment to be instantiated and executed with a single command.

The strength of \pkg{Docker} lies not only in its ability to encapsulate
the correct versions of \proglang{R} and its packages but also in its
inclusion of system-level dependencies. Since \pkg{Docker} images are
nearly complete Linux systems, they naturally bundle libraries and tools
required by certain R packages (for example, \pkg{nloptr}'s dependency
on the \proglang{nlopt} library). This ensures that future replicators
of a study have access to the same software environment, including all
necessary system components. Moreover, these images can be easily
shared, enabling seamless reproducibility across machines and
collaborators.

The Rocker project, introduced by \citet{boettiger2017}, provides a
collection of pre-built \pkg{Docker} images for the \proglang{R}
community. These images come with specific R versions and, in some
cases, preinstalled packages, making them convenient base images for
building reproducible analysis environments without starting from
scratch.

Despite its strengths, \pkg{Docker} can be cumbersome for interactive
use. Although containers can be modified at runtime, any changes are
lost once the container stops unless they are explicitly saved. Running
graphical applications from containers is also possible, but it is
generally difficult to configure and tends to work reliably only on
Linux systems. For this reason, \pkg{Docker} is most often used to run
web-based applications (such as integrated development environments,
IDEs, like the web version of RStudio) inside containers, offering a
reproducible yet flexible environment for interactive work.

Some IDEs provide built-in support for executing code directly within a
running container, allowing a more seamless workflow. However, these
setups typically require additional configuration, and many
general-purpose IDEs are not well optimized for statistical or data
science workflows (though Positron may soon change that).
